---
title: "Resequencing Analysis: Variant Annotation, Filtering & Prioritization"
author: "Mariangela Santorsola"
date: "2026/27"
output:
 html_document:
   toc: true
params:
    vcf:  /PATH/TO/joint_germline_recalibrated_snpEff.ann.vcf.gz
    ext_funcs:  /PATH/TO/extract_annotations_full.R
---
### Variant calling and annotation
```{r variant_settings, echo=FALSE, include=FALSE}
library(knitr)
library(VariantAnnotation)
library(tidyverse)
source(params$ext_funcs)
```

Read vcf file from nf-core/sarek:
```{r import_VCF, eval=TRUE, echo=TRUE, include=TRUE, warning=FALSE, message=FALSE}
vcf <- readVcf(params$vcf)
```


```{r rowRanges, eval=TRUE, echo=FALSE, include=FALSE}
head(rowRanges(vcf))
```

```{r infoSection, eval=TRUE, echo=FALSE, include=FALSE}
head(info(vcf))
```


```{r genoSection, eval=TRUE, echo=FALSE, include=FALSE}
head(geno(vcf)$GT)
```


Create the table of annotations by the INFO field within the VCF file:

```{r simplifyVCF, message=FALSE, warning=FALSE}
variants <- rowRanges(vcf)
variants$`REF` <- as.character(variants$`REF`)
variants$ALT <- sapply(variants$ALT, function(x) as.character(x)[1])
variants <- as_tibble(variants)
variants$variantName = names(rowRanges(vcf))
variants = cbind(variants, as_tibble(geno(vcf)$GT))
names(variants)[names(variants) == "patient_01_sample_01"] <- "case_case1"
names(variants)[names(variants) == "patient_02_sample_02"] <- "control_control1"
variants$gene <- unlist(lapply(info(vcf)$ANN, get_most_severe_gene))
variants$ens <- unlist(lapply(info(vcf)$ANN, get_most_severe_ens))
variants$transcript <- unlist(lapply(info(vcf)$ANN, get_most_severe_tr))
variants$aa_change <- unlist(lapply(info(vcf)$ANN, get_most_severe_aa_change))
variants$consequence <- unlist(lapply(info(vcf)$ANN, get_most_severe_consequence))
variants$impact <- unlist(lapply(info(vcf)$ANN, get_most_severe_impact))
```

### Localization and biological consequences of identified mutations


```{r consequences, eval=TRUE, echo=FALSE, include=TRUE, fig.cap="Distribution of the percentage of identified mutations across biological consequence categories"}
variants %>%
  filter(!is.na(consequence)) %>%
  count(consequence) %>%
  mutate(count = n,
         percent = paste0(round(count/sum(count) * 100, digits = 2), "%"))%>%
  arrange(desc(count)) %>%
  mutate(lab.ypos = cumsum(count) - 0.5 * count) %>%
  ggplot(aes(x="", y=count, fill=consequence))+
  geom_bar(width = 1, stat = "identity")+
  coord_polar(theta = "y")+
  ggrepel::geom_text_repel(aes(y=lab.ypos, label=percent), max.overlaps = Inf)+
  theme_void()
```
 
 or you can use a table format:
 
```{r consequences_table, eval=TRUE, echo=FALSE, include=TRUE}
variants %>%
  filter(!is.na(consequence)) %>%
  group_by(consequence) %>%
  summarise(count=n()) %>%
  arrange(desc(count)) %>%
  kable()
```


## Variant Filtering
### Filtering Criteria

The following *r* code block outputs the filtered variants based on the defined criteria:

```{r filterVariants, eval=TRUE, echo=TRUE, include=TRUE}
filteredVars <- variants %>%
  filter(
    control_control1 %in% c("0/0", "0|0") &               # accept both phased and unphased reference genotypes
    impact %in% c("MODERATE", "HIGH") &
    # ens %in% significant_de_genes &                   # functional impact filter
    consequence == "missense_variant"                    # biologically relevant variant type
  )

filteredVars <- as_tibble(filteredVars)
```

### Filtered mutations for further validation assays

```{r filtered_vars_table, eval=TRUE, echo=FALSE, include=TRUE}
selectedVars <- filteredVars %>%
  dplyr::select(seqnames, start, REF, ALT, gene, consequence, aa_change) 
selectedVars_table <- selectedVars 
selectedVars_table %>% knitr::kable()

```


```{r create_table_for_VEP, eval=TRUE, echo=TRUE, include=TRUE}
# create table for VEP
VEP.tab <- filteredVars %>%
  dplyr::select(seqnames, start, REF, ALT) %>%
  mutate(ID = ".",  .after = start) %>%
  mutate(seqnames = substr(seqnames, 4, nchar(.))) # all are from chromosome 21 and VEP does not accept "chr"
  
write.table(VEP.tab, "variants_to_VEP.txt", col.names = F, row.names = F, quote = FALSE)
```


### Exercise: Enriching variant annotations with VEP

Now it’s your turn! Ensembl VEP outputs rich functional annotations—including **AlphaMissense** pathogenicity predictions for missense variants.

**Your Challenge:**
Import your VEP output table into R, **add to your variants the additional relevant annotations from VEP** (e.g., SIFT, PolyPhen, AlphaMissense class/score, LOEUF), filter for prioritized variants, and include the final result table in your output.

