---
title: "Differential Gene Expression Analysis (DEA) - reporting: datasets_LABOS_exercise1"
author: "Mariangela Santorsola"
date: "2024/25"
output:
 html_document:
   toc: true
params:
    quant_path: /home/rstudio/salmon
    tx2gene: /home/rstudio/gencode.v29.transcripts_no-vers_chr21_tx2gene.txt
    
---

## Differential expression analysis

The goal of differential expression analysis is to compare gene expression levels between distinct experimental conditions, such as treatments and controls.

Determine which genes show statistically significant differences in expression between conditions.

Measure the magnitude and direction of changes in gene expression (e.g., upregulation or downregulation).

Understand the biological implications of expression changes, such as pathways, processes, or mechanisms influenced by treatments or conditions.


### Input data

**Note** Please double-check the file paths for all inputs referenced in this document.

- quant_path: Ensure that this params points to the directory "salmon" containing all individual sample folders:
```
.salmon                               # "salmon" directory created by nf-core/rnaseq 
├── **control_rep1**                  # Directory for an individual sample in cases/controls (e.g. cases, controls)
│   ├── aux_info
│   │   ├── ambig_info.tsv
│   │   ├── expected_bias.gz
│   │   ├── fld.gz
│   │   ├── meta_info.json
│   │   ├── observed_bias.gz
│   │   └── observed_bias_3p.gz
│   ├── cmd_info.json
│   ├── libParams
│   │   └── flenDist.txt
│   ├── lib_format_counts.json
│   ├── logs
│   │   └── salmon_quant.log
│   ├── quant.genes.sf
│   └── **quant.sf**
[...]
```
You can upload the compressed Salmon directory (e.g., salmon.zip) directly. RStudio will automatically uncompress it upon upload.

- tx2gene: ensure that you use the following file:: [gencode.v29.transcripts_no-vers_chr21_tx2gene.txt](https://github.com/santorsola-teaching/class-lab-adv-omics/blob/main/Tutoring_class_recap/gencode.v29.transcripts_no-vers_chr21_tx2gene.txt).


## R env setup

Load R libraries

```{r packages, eval=TRUE, echo=FALSE}
library(DESeq2)
library(tximport)
library(tidyverse)
library(pheatmap)
library(clusterProfiler)
library(DOSE)
library(org.Hs.eg.db)
```


## Set conditions, e.g. cases/controls

```{r}
dataset <- tibble(
  sample = c("control_rep1",
             "control_rep2",
             "control_rep3",
             "treatment_rep1",
             "treatment_rep2",
             "treatment_rep3"),
  condition = c(rep("control", 3),
                rep("case", 3))
)

dataset
```



## Read the table linking transcripts to genes 

```{r}
tx2gene <- read_tsv(params$tx2gene, col_names = TRUE, show_col_types = FALSE)

#tx2gene[tx2gene$TXNAME == "ENST00000624081",] #ENSG00000279493
```



## Read quantification files (*.quant) from salmon pseudo-alignment

```{r}
files <- file.path(params$quant_path, dataset$sample, "quant.sf")
names(files) <- dataset$sample

#quant <- read_tsv(files[1]) ## transcript-level data
#head(quant)
#quant[quant$Name == "ENST00000624081",]
```

quant.sf includes:

- Name: target transcript ID
- Length: length of the target transcript 
- EffectiveLength: computed effective length of the target transcript, based on all factors that will effect the probability of sampling fragments from this transcript, e.g. gc-content   
- TPM: transcripts-per-million, relative abundance measure, (exon reads in gene) / (total exon reads) x 1 million
- NumReads: salmon’s estimation of the number of reads



The function _tximport_ imports transcript-level estimates from various external software and optionally summarizes abundances, counts, and transcript lengths to the gene-level for DESeq2:

```{r}
txi <- tximport(files, type = "salmon", tx2gene = tx2gene)
colnames(txi$counts)
rownames(dataset) <- colnames(txi$counts)


#ls(txi)
#head(txi$abundance) ## provided by the quantification tools as TPM (transcripts-per-million)
#head(txi$counts)    ## estimated counts
#head(txi$length)    ## the effective gene lengths
#head(txi$countsFromAbundance) ## "no" default: not scaling
```


##  Construct a DESeqDataSet from the txi object and sample information in dataset

```{r}
dds <- DESeqDataSetFromTximport(txi, dataset, ~condition)


#x <- assays(dds)$counts
#x[rownames(x) == 'ENSG00000141956',]
```



## Prefiltering
Filter low count genes before running the DESeq2 functions, to reduce the memory size of the dds data object, and increase the speed of count modeling within DESeq2 

```{r}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```



## Set base level to "control"

```{r}
dds$condition <- relevel(dds$condition, ref = "control")
```



## Perform Differential Expression Analysis between cases and controls

```{r}
dds <- DESeq(dds)
```

DESeq2 main steps:

- estimating size factors: related to the library size (total number of reads in the library), considers that the most highly expressed genes take up the majority of the reads
- using 'avgTxLength' from assays(dds), correcting for library size
- estimating dispersions
- gene-wise dispersion estimates
- mean-dispersion relationship
- final dispersion estimates
- fitting model and testing
## fit curve to gene-wise dispersion estimates
## shrink gene-wise dispersion estimates toward the values predicted by the curve


##  Extract results table from the dds object

```{r}
res <- results(dds)
head(results(dds, tidy=TRUE)) #let's look at the results table 
```

By default DESeq2 uses the Wald test to identify genes that are differentially expressed between two sample classes.
To get information on each field in results:

```{r}
mcols(res, use.names=TRUE)
```


## DE analysis summary

```{r}
summary(res)
```



### Order Genes by padj

```{r}
resOrdered <- res[order(res$padj),]
resOrdered
```


### Data visualization

#### MA plot
The MA plot shows the mean of the normalized counts versus the log2 foldchanges for all genes tested. The genes that are significantly DE are colored (blue) to be easily identified.

```{r maplot, eval=TRUE, echo=FALSE, include=TRUE, fig.cap="Plot showing the log2 fold change over the mean of normalized counts for analysed samples"}
plotMA(res, ylim=c(-3,3))
#abline(h=c(-1,1), col="dodgerblue", lwd=2)
```



## Dispersion plot

```{r}
#pdf("plots_dispersion.pdf")
plotDispEsts(dds)
#dev.off()
```




# Plot counts 

Useful to inspect the counts of reads for a single gene across the groups. 
```{r}
#pdf("plots_counts.pdf")
plotCounts(dds, gene=which.min(res$padj), intgroup="condition") # e.g. gene with the lowest padj value
#dev.off() 

```



## Save results to file or RData

```{r}
resdata <- as_tibble(resOrdered)
resdata$gene <- rownames(resOrdered)
write_tsv(resdata, "deseq2_results.tsv")
save.image("deseq2_analysis.RData")
```



## CLUSTERING analysis

```{r}
ntd <- normTransform(dds)        
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]

df <- as.data.frame(colData(dds)[,c("condition")])
```



### CLUSTERING visualization

Useful to explore the count matrix.

```{r}
#pdf("plots_heatmap.pdf")
pheatmap(assay(ntd)[select,],
         cluster_cols=FALSE, annotation_col=df$condition)
#dev.off()
```



## Plot Principal Component Aanalysis
Useful for visualizing the overall effect of experimental covariates and batch effects.

```{r}
#pdf("plots_pca.pdf")
plotPCA(ntd, intgroup=c("condition"))
#dev.off()
```

### List of Differentially Expressed Genes

```{r significant_genes_table, eval=TRUE, echo=FALSE, include=TRUE}
resOrdered <- res[order(res$padj),]
resOrdered <- na.omit(resOrdered) ### omit "NA" values 
resOrdered$Gene <- rownames(resOrdered)
sign_de_genes_table <- as_tibble(resOrdered[resOrdered$padj < 0.1,])
sign_de_genes_table %>% 
    dplyr::select(Gene, log2FoldChange, pvalue, padj) %>%
    knitr::kable()
significant_de_genes <- sign_de_genes_table$Gene
```



```{r}
### check gene Ensembl ID / Gene symbol
#sig_genes_names <- tx2gene[which(tx2gene$gene_id %in% sig_genes),]
```


## Gene Ontology Enrichment Analysis


```{r GO_BP}
# ego <- enrichGO( gene = significant_de_genes, # Foreground genes: DEGs
#                  universe = unique(tx2gene$GENEID), # Background: all measured genes
#                  OrgDb = org.Hs.eg.db, # Annotations for Human genes
#                  keyType = 'ENSEMBL',
#                  ont = "BP", # The GO three main aspects are: Molecular Function (MF), Cellular Component (CC) and Biological Process (BP)
#                  pAdjustMethod = "BH",
#                  pvalueCutoff  = 0.05)
# 
# ego
```



```{r GO_CC}
ego <- enrichGO( gene = significant_de_genes,
                 universe = unique(tx2gene$GENEID),
                 OrgDb = org.Hs.eg.db,
                 keyType = 'ENSEMBL',
                 ont = "CC", # The GO three main aspects are: Molecular Function (MF), Cellular Component (CC) and Biological Process (BP)
                 pAdjustMethod = "BH",
                 pvalueCutoff  = 0.05)

ego
```


#### Visualise enriched GO-CC terms

```{r}
p <- dotplot(ego, showCategory=20) + ggtitle("Top 20 Enriched Biological Processes")
p
```



```{r GO_MF}
# ego <- enrichGO( gene = significant_de_genes,
#                  universe = unique(tx2gene$GENEID),
#                  OrgDb = org.Hs.eg.db,
#                  keyType = 'ENSEMBL',
#                  ont = "MF", # The GO three main aspects are: Molecular Function (MF), Cellular Component (CC) and Biological Process (BP)
#                  pAdjustMethod = "BH",
#                  pvalueCutoff  = 0.05)
# 
# ego

```


#### Gene-Concept Network

```{r}
cnetplot(ego, foldChange=resOrdered$log2FoldChange[which(resOrdered$padj<0.5)])

```


